---
title: "Part 1: Simulation using the exponential distribution"
author: "T. Mashburn"
date: "1-24-2023""
output:
    pdf_document:
        highlight: zenburn
fontsize: 10pt
geometry: margin=0.8in
---
*Project for the "Statistical Inference" course (Coursera, Aug. 2014)*

### Comparing the simulated mean and variance with the theoretical values

We will run 1000 rounds of simulation of 40 exponentials with $\lambda = 0.2$,
using a fixed seed, and comparing the distribution of the simulated mean
and variance with the theoretical value of $1 / \lambda$:

```{r results='asis'}
library(pander)
nsim <- 1000
nvals <- 40
lambda <- 0.2
set.seed(567)
simdata <- t(replicate(nsim, rexp(nvals, lambda)))
df <- data.frame(Mean=c(mean(rowMeans(simdata)), 1/lambda),
                 Variance=c(mean(apply(simdata, 1, var)), 1/lambda^2))
rownames(df) <- c("Simulated", "Theoretical")
pander(df, round=2)
```

The simulated and theoretical values are very close, as expected by the CLT.

### Assessing if the simulated values are approximately normal

Also, according to the CLT, the distribution of the simulated means 
should be approximately normal. To illustrate this we will
normalize the vectors and compare it to a $N(0,1)$ distribution.

```{r}
library(ggplot2)
meanvals <- rowMeans(simdata)
zmean <- (meanvals - mean(meanvals)) / sd(meanvals)
qplot(zmean, geom = "blank") +
    geom_line(aes(y = ..density.., colour = 'Empirical'), stat = 'density') +  
    stat_function(fun = dnorm, aes(colour = 'Normal')) +                       
    geom_histogram(aes(y = ..density..), alpha = 0.4, binwidth=.35) +
    geom_vline(xintercept=0, colour="red", linetype="longdash") +
    scale_colour_manual(name = 'Density', values = c('red', 'blue')) +
    ylab("Density") + xlab("z") + ggtitle("Mean values distribution") +
    theme_bw() + theme(legend.position = c(0.85, 0.85))
```

### Evaluating the coverage of the confidence interval

Theoretically, a 95% confidence interval should contain, if we simulate
a big number of them, the mean value for the exponential distribution
($1 / \lambda$) 95% of the time.

```{r}
set.seed(567)
lambda <- 0.2
# checks for each simulation if the mean is in the confidence interval
inconfint <- function(lambda) {
                ehats <- rexp(1000, lambda)
                se <- sd(ehats)/sqrt(1000)
                ll <- mean(ehats) - 1.96 * se
                ul <- mean(ehats) + 1.96 * se
                (ll < 1/lambda & ul > 1/lambda)
        }
# estimate the coverage in each round of simulations
coverage <- function(lambda) {
    covvals <- replicate(100, inconfint(lambda))
    mean(covvals)
}
# perform the simulation
simres <- replicate(100, coverage(lambda))
mean(simres)
```

As expected, the confidence interval contains the theoretical value 94.84% of the
time (close to the expected 95%).
statinference-assignment-part2.Rmd
---
title: "Part 2: Analysis of the ToothGrowth data set"
author: "Jesus M. Castagnetto"
date: "2014-08-23"
output:
    pdf_document:
        highlight: zenburn
fontsize: 10pt
geometry: margin=0.8in
---

*Project for the "Statistical Inference" course (Coursera, Aug. 2014)*

The `ToothGrowth` data set contains "... *The response is the length of
odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of 
Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice 
or ascorbic acid).* ..." [Ref.; C. I. Bliss (1952) The Statistics of Bioassay.
Academic Press] 

### Quick exploration of the data set

For the purposes of this analysis, we will convert the dose variable into a factor.

```{r}
data(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
```

The distributions of tooth length by delivery method and dose levels (see graph
in the next page), seem to indicate that for 0.5 and 1 mg, the mean tooth length
is greater for those where orange juice was used. For the 2 mg
level, no noticeable difference in the mean is observed, but the
dispersion for the OJ group seems to be smaller. 

### Multiple hypothesis testing

To ascertain whether there is a real difference between the groups by
dose level and delivery method, we will employ a series of two-sided
unpaired t-tests to obtain the confidence intervals and p-values.

The comparison will be done at each dose level, between delivery methods,
asumming unequal variance (supported by the distribution graph). The p-values will
be adjusted using the (conservative) Bonferroni correction, and the comparative
results shown in the table below.

Our null hypothesis in all cases is that there is no difference in the means
between the two groups (two-sided test).

```{r results='asis'}
library(pander)
ts <- lapply(c(.5, 1, 2), function(x) {
    t.test(len ~ supp, data=subset(ToothGrowth, dose==x), paired=FALSE, var.equal=FALSE)
    })
pvals <- c(ts[[1]]$p.value, ts[[2]]$p.value, ts[[3]]$p.value)
stats <- c(ts[[1]]$statistic, ts[[2]]$statistic, ts[[3]]$statistic)
adjp <- p.adjust(pvals, method = "bonferroni")
lls <- sapply(c(ts[[1]]$conf.int[1], ts[[2]]$conf.int[1], ts[[3]]$conf.int[1]), round, 3)
uls <- sapply(c(ts[[1]]$conf.int[2], ts[[2]]$conf.int[2], ts[[3]]$conf.int[2]), round, 3)
df <- data.frame(dose=c(0.5, 1, 2), t=stats, p=pvals, adj=adjp,
                 ci=paste0("[",paste(lls, uls, sep=", "), "]"))
colnames(df) <- c("Dose", "t", "p-value", "adj. p-value", "conf. int.")
pander(df, round=3, split.tables=120,
       caption="*Two-sided comparison of delivery methods by dose*")
```

```{r}
library(ggplot2)
ggplot(data=ToothGrowth, aes(y=len, x=supp, fill=supp)) + geom_boxplot() +
    facet_wrap(~ dose, ncol=3) + ylab("Tooth length") + xlab("Delivery method") + 
    ggtitle("Tooth growth by delivery and dose") + 
    stat_summary(fun.y=mean, geom="point", shape=5, size=2) +
    theme_bw()+ theme(legend.position = "none")
```

### Conclusions

- At the 0.5 and 1 mg dose levels, there is a statistically significant difference
  between the means of the OJ and VC groups. The adjusted p-values are significant
  at the $\alpha$ = 0.05 level, and the 95% confidence intervals do not include zero.
- For the 2 mg dose level, we fail to reject the null hypothesis, the adjusted
  p-value is much greater than 0.5, and the 95% confidence interval includes zero.
  So, it seems that at that Vitamin C level, there is no significative influence
  of the delivery method on tooth growth in guinea pigs.
- Because the effect size is very small for the 2 mg level, to be able to detect
  a significative difference, we would need a much bigger sample size (n=10 does
  not give the test enough power)
```{r}
s <- subset(ToothGrowth, dose==2)
d <- split(s, s$supp)
n1 <- 10; m1 <- mean(d$OJ$len); s1 <- sd(d$OJ$len)
n2 <- 10; m2 <- mean(d$VC$len); s2 <- sd(d$VC$len)
pooled_sd <-  sqrt( ((n1 - 1) * s1^2 + (n2-1) * s1^2) / (n1 + n2-2))
cat("effect size:", round((m2 - m1)/pooled_sd, 3), "\nestimated sample size:",
    round(power.t.test(power=0.9, delta = m1 - m2, sd=pooled_sd)$n,0))
```


